Circumnuclear Media and Accetion Rates of Quiescent Supermassive Black Holes

### 2015-02-10 ms.tex
more edits 6.3

Radiative cooling usually has its greatest impact near or external to the stagnation radius, where the gas resides in near hydrostatic balance.  If radiative cooling becomes important, it can qualitatively alter key features of the accretion flow. Initially hydrostatic gas that cools at constant pressure $P \propto nT$ at the rate $\dot{q}_{\rm rad} = n^{2}\Lambda \propto T^{-2.7}$ (eq.~[\ref{eq:qdot_rad}]) is thermally unstable, where $\Lambda(T) \propto T^{-0.7}$ (see below).  Thermal instability can lead to enhanced star formation and temporarily increases the black hole accretion rate; when coupled to feed-back processes, the latter can lead to time-dependent (e.g. limit cycle) behavior which is inconsistent with our assumption of steady flow (e.g.~\citealt{Ciotti&Ostriker07}; \citealt{Ciotti+10}; \citealt{Gan+14}).

@@

Feedback from accretion onto the SMBH represents an important source of heating which, however, is also the most difficult to quantify.  A key difference between AGN heating and the other sources discussed thus far is its dependence on the SMBH accretion rate $\dot{M}$, which is itself a function of the heating rate (eq.~[\ref{eq:mdot_analytic}]).

There are two types of SMBH feedback: kinetic and radiative.  Radiative feedback is potentially effective even in low luminosity AGN via Compton heating, which provides a volumetric heating rate (\citealt{Gan+14})
\be
\dot{e}_{\rm C} = 4.1\times 10^{-35}n^{2}\xi T_{\rm C}\,{\rm erg\,cm^{-3}\,s^{-1}},
\ee
where $\xi = L/n r^{2}$ is the ionization parameter and $L$ is the SMBH luminosity with Compton temperature $T_{\rm C} \sim 10^{9}$ K $\gg T$ (\citealt{Gan+14}).

{\bf AG: Note that the Compton heating will depend on the distance of
  the hole as $1/r^2$ so Compton heating will be stronger closer to the
  black hole! }
The importance of Compton heating can be estimated by assuming the SMBH
radiates with a luminosity $L = \epsilon_{\rm rad} \dot{M}c^{2}$, where
$\epsilon_{\rm rad}$ is the radiative efficiency and where $\dot{M}$ is
estimated from equation (\ref{eq:mdot_analytic}).  Then using
equations (\ref{eq:rs_simple}), (\ref{eq:mdot_analytic}),
(\ref{eq:rhors}), (\ref{eq:rhostarrs}) we calculate that from equation
(\ref{eq:vw_eff}) that the effective Compton heating rate at the stagnation radius is given by
\begin{align} v_{w}^{\rm C} \simeq
  \begin{cases} 29 \eta_{0.02}^{0.5}T_{\rm
C,9}^{0.5}\epsilon_{-2}^{0.5} M_{\bullet,8}^{0.38}v_{500}^{-1.4}\,{\rm
km\,s^{-1}} &, \text{core}\\ 50 \eta_{0.02}^{0.5}T_{\rm
C,9}^{0.5}\epsilon_{-2}^{0.5} M_{\bullet,8}^{0.24}v_{500}^{-0.7}\,{\rm
km\,s^{-1}} &, \text{cusp},
  \end{cases}
  \label{eq:vC}
\end{align} where $T_{C,9} = T_{C}/10^{9}$ K and $\epsilon_{-2} =
\epsilon_{\rm rad}/0.01 \sim 1$ is the maximum radiative efficiency
for an advective accretion flow expected to produce high $T_{\rm C}$
emission (e.g.~\citealt{Narayan&Yi95}).  If heating is provided by
Compton heating alone, i.e. $\tilde{v}_{w} = v_{w}^{C}$, then equation
(\ref{eq:vC}) can be solved for $\tilde{v}_{\rm w}$ to find

@@

Kinetic feedback occurs from an outflow of mass or energy from the
viscinity of the black hole in the form of a disk wind or jet, which
deposits its energy across a much wider range of radii.  If the total
outflow power is proportional to the SMBH accretion rate, $L_{\rm j} =
\epsilon_{\rm j} \dot{M}c^{2}$, where $\epsilon_{\rm j} < 1$ is an
efficiency factor.  Assuming that this energy is deposited uniformly
out to a radius $r_{\rm heat}$ and volume $V_{\rm heat} \propto r_{\rm
heat}^{3}$, then the volumetric heating rate $e = \epsilon_{\rm
j}\dot{M}c^{2}/V_{\rm heat}$ results in an effective wind velocity at
the stagnation radius given by

@@

\begin{itemize}
\item{Fact that some low mass BHs appear to have nearly Eddington accretion rates, suggest an external origin - since the maximum Ia rate appears to be below this}
\end{itemize}

>>>>

Radiative cooling usually has its greatest impact near or external to the stagnation radius, where the gas resides in near hydrostatic balance.  If radiative cooling becomes important, it can qualitatively alter key features of the accretion flow. Initially hydrostatic gas that cools at constant pressure $P \propto nT$ at the rate $\dot{q}_{\rm rad} = n^{2}\Lambda \propto T^{-2.7}$ (eq.~[\ref{eq:qdot_rad}]) is thermally unstable, where $\Lambda(T) \propto T^{-0.7}$ (see below).  Thermal instability can lead to enhanced star formation and temporarily increases the black hole accretion rate; when coupled to feed-back processes, the latter can lead to time-dependent (e.g. limit cycle) behavior which is inconsistent with our assumption of steady flow (e.g.~\citealt{Ciotti&Ostriker07}; \citealt{Ciotti+10}; \citealt{Yuan&Li11}, \citealt{Gan+14}).

@@

Feedback from accretion onto the SMBH represents an important source of heating which, however, is also the most difficult to quantify (e.g., \citealt{Brighenti&Mathews03}, \citealt{DiMatteo+05}; \citealt{Kurosawa&Proga09}; \citealt{Fabian12} for a recent review).  A key difference between AGN heating and the other sources discussed thus far is its dependence on the SMBH accretion rate $\dot{M}$, which is itself a function of the heating rate (eq.~[\ref{eq:mdot_analytic}]).

There are two types of SMBH feedback: kinetic and radiative.  Radiative feedback is potentially effective even in low luminosity AGN via Compton heating (e.g., \citealt{Sazonov+04}, \citealt{Ciotti+10}), which provides a volumetric heating rate (\citealt{Gan+14})
\be
\dot{e}_{\rm C} = 4.1\times 10^{-35}n^{2}\xi T_{\rm C}\,{\rm erg\,cm^{-3}\,s^{-1}},
\ee
where $\xi = L/n r^{2}$ is the ionization parameter and $L$ is the SMBH luminosity with Compton temperature $T_{\rm C} \sim 10^{9}$ K $\gg T$ (e.g., \citealt{Ho99}, \citealt{Eracleous+10}).

The importance of Compton heating can be estimated by assuming the SMBH
radiates with a luminosity $L = \epsilon_{\rm rad} \dot{M}c^{2}$, where
$\epsilon_{\rm rad}$ is the radiative efficiency and where $\dot{M}$ is
estimated from equation (\ref{eq:mdot_analytic}).  Then using
equations (\ref{eq:rs_simple}), (\ref{eq:mdot_analytic}),
(\ref{eq:rhors}), (\ref{eq:rhostarrs}) we calculate that from equation
(\ref{eq:vw_eff}) that the effective Compton heating rate at the stagnation radius is given by
\begin{align} v_{w}^{\rm C} \simeq
  \begin{cases} 29 \eta_{0.02}^{0.5}T_{\rm
C,9}^{0.5}\epsilon_{-2}^{0.5} M_{\bullet,8}^{0.38}v_{500}^{-1.4}\,{\rm
km\,s^{-1}} &, \text{core}\\ 50 \eta_{0.02}^{0.5}T_{\rm
C,9}^{0.5}\epsilon_{-2}^{0.5} M_{\bullet,8}^{0.24}v_{500}^{-0.7}\,{\rm
km\,s^{-1}} &, \text{cusp},
  \end{cases}
  \label{eq:vC}
\end{align} where $T_{C,9} = T_{C}/10^{9}$ K and $\epsilon_{-2} =
\epsilon_{\rm rad}/0.01 \sim 1$ is the maximum radiative efficiency
for an advective accretion flow expected to produce high $T_{\rm C}$
emission (e.g.~\citealt{Narayan&Yi95}).  If heating is provided by
Compton heating alone, i.e. $\tilde{v}_{w} = v_{w}^{C}$, then equation
(\ref{eq:vC}) can be solved for $\tilde{v}_{\rm w}$ to find

@@

Kinetic feedback occurs from an outflow of mass or energy from the
viscinity of the black hole in the form of a disk wind or jet, which
deposits its energy across a much wider range of radii (e.g.~\citealt{McNamara&Nulsen07}; \citealt{Novak+11}; \citealt{Gaspari+12}).  If the total
outflow power is proportional to the SMBH accretion rate, $L_{\rm j} =
\epsilon_{\rm j} \dot{M}c^{2}$, where $\epsilon_{\rm j} < 1$ is an
efficiency factor.  Assuming that this energy is deposited uniformly
out to a radius $r_{\rm heat}$ and volume $V_{\rm heat} \propto r_{\rm
heat}^{3}$, then the volumetric heating rate $e = \epsilon_{\rm
j}\dot{M}c^{2}/V_{\rm heat}$ results in an effective wind velocity at
the stagnation radius given by

@@

\begin{itemize}
\item{Fact that some low mass BHs appear to have nearly Eddington accretion rates, suggest an external origin - since the maximum Ia rate appears to be below this}
\item{Growth of low mass BHS, e.g. \citealt{Kauffmann&Heckman09}}
\end{itemize}
